---
title: "1.1 Exploratory Data Analysis (EDA)"
author: "Thoa Le (a1847933)"
output: pdf_document
---

Data from 1960 to 2024 on Total Live Births (TLB) and Total Fertility Rates (TFR) was pulled from Singapore Department of Statistics (data.gov.sg). The data was then transformed into a tsibble object with three columns: Year, Total Fertility Rates (TFR) and Total Live Births.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
library(fpp3) # time series analysis package
library(tsibble) # tsibble objects
library(lubridate) # for data to work with dates and times
library(feasts)
library(latex2exp)
library(patchwork) # putting 2 plots next to each other

data <- readr::read_csv("BirthsAndFertilityRatesAnnual.csv") # read in the data
BirthAndFertility <- data |>
  pivot_longer( # change rows to column
    cols = `1960`:`2025`,
    names_to = "Year",
    values_to = "Value", # naming it to remember that it needs to be values
    values_transform = list(Value = as.numeric)
  ) |>
  filter(DataSeries %in% c("Total Fertility Rate (TFR)", "Total Live-Births")) |>
  mutate(Year = as.integer(Year)) |>
  filter(Year != 2025) |> # assignment specified until 2024
  pivot_wider( # taking TFR & Total Live-Births into 2 separate columns according to year
  names_from = DataSeries, 
  values_from = Value
  )
BirthAndFertility <- as_tsibble(BirthAndFertility, index = Year) # created a tsibble
is_tsibble(BirthAndFertility) # checking it is a tsibble
```

In this exploratory data analysis, the data was visualised to identify patterns and trends over time. Since the data is yearly, classical decomposition and the X-11 method are not applicable. Once the patterns are identified, appropriate models can be selected, fitted, and evaluated for forecasting.

\newpage
```{r page-2, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility, `Total Fertility Rate (TFR)`) +
  labs(title = "Initial Plot of Total Fertility Rate (TFR)")
autoplot(BirthAndFertility,`Total Live-Births`)+
  labs(title = "Initial Plot of Total Live-Births")
```

From the initial plots of TFR and TLB, both series show a decreasing trend over time, with TFR exhibiting a more pronounced decline. As the data is yearly, seasonality cannot be determined. While both series may suggest cyclic   behaviour, the trend would need to be removed to confirm this.

\newpage
```{r page-3}
BirthAndFertility |>
  ggplot(aes(x = `Total Fertility Rate (TFR)`, y = `Total Live-Births`)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Total Fertility Rate (TFR) vs Total Live-Births (TLB)")
```

The relationship between TFR and TLB was examined by plotting the two variables against each other. There appears to be a nonlinear positive relationship, suggesting that higher fertility rates are associated with higher numbers of live births, although further analysis is required.

\newpage
```{r page-4, echo = TRUE, warning = FALSE, message = FALSE}
p1 <- BirthAndFertility |>
  gg_lag(`Total Fertility Rate (TFR)`, geom = 'point') +
  labs(title = "Lag Plot of Total Fertility Rate (TFR)")
p2 <- BirthAndFertility |>
  gg_lag(`Total Live-Births`, geom = 'point') +
  labs(title = "Lag Plot of Total Live-Births (TLB)")
p1 + p2 # placing the plots side by side
```

The above lag plots for TFR and TLB indicate stronger positive relationships at lags 1 and 2 suggesting that AR(1) and AR(2) models could be appropriate.

The following fitted AR(1) and AR(2) models (orange lines) follow the TFR and TLB (grey lines) but tends to overshoots and underestimate peaks. Residuals were examined and appear to have no pattern; however the ACF plots show multiple spikes outside the significance bounds indicating the residuals are not white noise. This indicates that an ARIMA model, which can account for trend, may be more appropriate.

\newpage
```{r page-5, out.width = "70%", fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE}
TFR_ar <- BirthAndFertility |> model(AR(`Total Fertility Rate (TFR)`, order = 1)) # fits AR model
TFR_fits <- augment(TFR_ar) # adds columns of data
ggplot(TFR_fits, aes(x = Year)) +
  geom_line(aes(y =`Total Fertility Rate (TFR)`), colour = "grey") + # raw data
  geom_line(aes(y = .fitted), colour = "#D55E00") + # 
  labs(title = "Plot of AR(1) fitted model (orange) and raw data (grey) for TFR")

TLB_ar <- BirthAndFertility |> model(AR(`Total Live-Births`, order = 1))
TLB_fits <- augment(TLB_ar)
ggplot(TLB_fits, aes(x = Year)) +
  geom_line(aes(y =`Total Live-Births`), colour = "grey") + # raw data
  geom_line(aes(y = .fitted), colour = "#D55E00") + # 
  labs(title = "Plot of AR(1) fitted model (orange) and raw data (grey) for TLB")
```

\newpage
```{r page-6, out.width = "70%", fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE}
TFR_ar2 <- BirthAndFertility |> model(AR(`Total Fertility Rate (TFR)`, order = 2)) # fits AR model
TFR_fits2 <- augment(TFR_ar2) # adds columns of data
ggplot(TFR_fits2, aes(x = Year)) +
  geom_line(aes(y =`Total Fertility Rate (TFR)`), colour = "grey") + # raw data
  geom_line(aes(y = .fitted), colour = "#D55E00") + # 
  labs(title = "Plot of AR(2) fitted model (orange) and raw data (grey) for TFR")

TLB_ar2 <- BirthAndFertility |> model(AR(`Total Live-Births`, order = 2))
TLB_fits2 <- augment(TLB_ar2)
ggplot(TLB_fits2, aes(x = Year)) +
  geom_line(aes(y =`Total Live-Births`), colour = "grey") + # raw data
  geom_line(aes(y = .fitted), colour = "#D55E00") + # 
  labs(title = "Plot of AR(2) fitted model (orange) and raw data (grey) for TLB")
```
\newpage
```{r page-7, out.width = "85%", fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE}
gg_tsresiduals(TFR_ar) + # check residuals
  labs(title = "AR(1) model Residuals for TFR")
gg_tsresiduals(TLB_ar) +
  labs(title = "AR(1) model Residuals for TLB")
```

\newpage
```{r page-8, out.width = "85%", fig.align = 'center', echo = TRUE, warning = FALSE, message = FALSE}
gg_tsresiduals(TFR_ar2) + # check residuals
  labs(title = "AR(2) model Residuals for TFR")
gg_tsresiduals(TLB_ar2) +
  labs(title = "AR(2) model Residuals for TLB")
```

\newpage
```{r page-9, out.width = "70%", fig.align = 'center'}
BirthAndFertility |>
  ACF(`Total Fertility Rate (TFR)`, lag_max = 9) |>
  autoplot() +
  labs(title = "ACF Plot for Total Fertility Rate (TFR)")
BirthAndFertility |>
  ACF(`Total Live-Births`, lag_max = 9) |>
  ggplot2::autoplot() +
  labs(title = "ACF Plot for Total Live-Births (TLB)")
```
The ACF for TFR shows significant positive autocorrelations from lags 1 to 9, while TLB shows significant positive autocorrelations from lags 1 to 6, with a slow decrease in the ACF as the lag increases. This suggests that neither series is white noise, as there are multiple spikes outside the significance bounds. This slow decrease indicates the presence of a trend and consistent with a non-stationary series. The absence of a sharp cut-off suggests that an autoregressive (AR) model may be suitable for this series.
```{r page-10}
BirthAndFertility |>
  ACF(`Total Fertility Rate (TFR)`, lag_max = 24) |>
  autoplot() +
  labs(title = "ACF Plot for Total Fertility Rate (TFR)")
```
An extended ACF plot for TFR confirms that significant positive autocorrelations persist up to lag 10.

\newpage
```{r page-11}
lambdatfr <- BirthAndFertility |>
  features(`Total Fertility Rate (TFR)`, features = guerrero) |> # Guerrero method
  pull(lambda_guerrero) # Box-Cox transformation parameter = -0.44514
BirthAndFertility |>
  autoplot(box_cox(`Total Fertility Rate (TFR)`, lambdatfr)) +
  labs(title = latex2exp::TeX(paste0(
    "Transformed Total Fertility Rate (TFR) with $\\lambda$ = ", round(lambdatfr,2))))
```
A Box-Cox transformation was considered, with the estimated lambda for TFR found to be below zero but approximately midway to -1, suggesting that a moderate transformation may be appropriate. The transformed series still shows a decreasing trend and there may also be cyclical behaviour; however the trend should be removed before confirming the presence of any cycles.

\newpage
```{r page-12}
lambdatlb <- BirthAndFertility |>
  features(`Total Live-Births`, features = guerrero) |> # Guerrero method
  pull(lambda_guerrero) # Box-Cox transformation parameter = -0.89992
BirthAndFertility |>
  autoplot(box_cox(`Total Live-Births`, lambdatlb)) +
  labs(title = latex2exp::TeX(paste0(
    "Transformed Total Live-Births (TLB) with $\\lambda$ = ", round(lambdatlb,2))))
```
The estimated lambda for TLB was also found to be below zero but closer to -1, suggesting that a strong transformation may be required. The decreasing trend from 1960 to 2024 appears less prominent in this plot, and there might also be cyclical behaviour; however, the trend should be removed before confirming the presence of any cycles.

\newpage
```{r}
dcmptfr <- BirthAndFertility |>
  model(stl = STL(`Total Fertility Rate (TFR)`))
components(dcmptfr) |>
  as_tsibble() |>
  autoplot(`Total Fertility Rate (TFR)`, colour = "grey") + # raw data
  geom_line(aes(y = trend), colour = "#D55E00") + # trend-cycle component
  labs(title = "Plot of trend-cycle component (orange) and raw data (grey) for TFR")
```
The STL decomposition suggests the series follows an additive structure with a clear trend component. 
# little variability to the raw data, more smooth

```{r}
dcmptlb <- BirthAndFertility |>
  model(stl = STL(`Total Live-Births`))
components(dcmptlb) |>
  as_tsibble() |>
  autoplot(`Total Live-Births`, colour = "grey") + # raw data
  geom_line(aes(y = trend), colour = "#D55E00") + # trend-cycle component
  labs(title = "Plot of trend-cycle component (orange) and raw data (grey) for TLB")
```
# more variability to the raw data, bigger changes

# STL decomposition
```{r}
components(dcmptfr) |> autoplot()
```
# decreasing trend as mentioned earlier

```{r}
components(dcmptlb) |> autoplot()
```
# overall decreasing trend but there was an increase from 1980 with a peak in 1991 before decreasing again

# the remainder of both decompositions resemble each other with sharp dip around 1986 and spike 1988

# moving average
```{r}
BirthAndFertility_MA_TFR <- BirthAndFertility |>
  mutate(`5-MA_TFR` = slider::slide_dbl(`Total Fertility Rate (TFR)`, mean,
                                  .before = 2, .after = 2, .complete = TRUE))
BirthAndFertility_MA_TFR |>
  autoplot(`Total Fertility Rate (TFR)`) +
  geom_line(aes(y = `5-MA_TFR`), colour = "#D55E00") +
  labs(title = "Total Fertility Rate (black) and 5-MA estimate of trend-cycle (orange)")
```
# the original data was already quite smooth so it matches quite closely with the trend-cycle

```{r}
BirthAndFertility_MA_TLB <- BirthAndFertility |>
  mutate(`5-MA_TLB` = slider::slide_dbl(`Total Live-Births`, mean,
                                  .before = 2, .after = 2, .complete = TRUE))
BirthAndFertility_MA_TLB |>
  autoplot(`Total Live-Births`) +
  geom_line(aes(y = `5-MA_TLB`), colour = "#D55E00") +
  labs(title = "Total Live-Births (black) and 5-MA estimate of trend-cycle (orange)")
```
# can see smooth curve but due to the roughness of the data should increase the average for a smoother curve

```{r}
BirthAndFertility_MA_TLB7 <- BirthAndFertility |>
  mutate(`7-MA_TLB` = slider::slide_dbl(`Total Live-Births`, mean,
                                  .before = 3, .after = 3, .complete = TRUE))
BirthAndFertility_MA_TLB7 |>
  autoplot(`Total Live-Births`) +
  geom_line(aes(y = `7-MA_TLB`), colour = "#D55E00") +
  labs(title = "Total Live-Births (black) and 7-MA estimate of trend-cycle (orange)")
```
# the curve is smoother at the cost of the ends

# evidence of the presence of business cycles?
# outliers that need to be explained by experts?
