---
title: "1.2 Final Report"
author: "Thoa Le (a1847933)"
output: pdf_document
---

# Abstract

# Introduction

# Methods
Data from 1960 to 2024 on Total Live Births (TLB) and Total Fertility Rates (TFR) was pulled from Singapore Department of Statistics (data.gov.sg). The data was then transformed into a tsibble object with three columns: Year, Total Fertility Rates (TFR) and Total Live Births.

```{r page-0, echo = FALSE, warning = FALSE, message = FALSE}
library(fpp3) # time series analysis package
library(tsibble) # tsibble objects
library(lubridate) # for data to work with dates and times
library(feasts)
library(forecast) # for arima
library(latex2exp)
library(patchwork) # putting 2 plots next to each other

data <- readr::read_csv("BirthsAndFertilityRatesAnnual.csv") # read in the data
BirthAndFertility <- data |>
  pivot_longer( # change rows to column
    cols = `1960`:`2025`,
    names_to = "Year",
    values_to = "Value", # naming it to remember that it needs to be values
    values_transform = list(Value = as.numeric)
  ) |>
  filter(DataSeries %in% c("Total Fertility Rate (TFR)", "Total Live-Births")) |>
  mutate(Year = as.integer(Year)) |>
  filter(Year != 2025) |> # assignment specified until 2024
  pivot_wider( # taking TFR & Total Live-Births into 2 separate columns according to year
  names_from = DataSeries, 
  values_from = Value
  )
BirthAndFertility <- as_tsibble(BirthAndFertility, index = Year) # created a tsibble
# is_tsibble(BirthAndFertility) # checking it is a tsibble

TFR <- BirthAndFertility |>
  filter(Year <= 2012) |>
  select(Year, `Total Fertility Rate (TFR)`) |>
  as_tsibble(index = Year)
# is_tsibble(TFR)

TLB <- BirthAndFertility |>
  filter(Year <= 2012) |>
  select(Year, `Total Live-Births`) |>
  as_tsibble(index = Year)
# is_tsibble(TLB)
```
From the exploratory data analysis, the data was visualised to identify patterns and trends over time. Since the data is yearly, classical decomposition and the X-11 method are not applicable and seasonal patterns cannot be detected. Additionally, the socio-economic and political factors influencing fertility are not included. Once the patterns are identified, appropriate models can be selected, fitted, and evaluated for forecasting.

\newpage
```{r page-1 TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility, `Total Fertility Rate (TFR)`) +
  labs(title = "Initial Plot of Total Fertility Rate (TFR)")
autoplot(TFR) +
  labs(title = "Initial Plot of Total Fertility Rate (TFR) until 2012")
```
From the initial plots of TFR and TLB, both series show a long-term decreasing trend over time, with TFR exhibiting a more pronounced decline. A notable change is observed around 1980 to 1990, where both series display a temporary broad peak before continuing their downward trend. As the data is yearly, seasonality cannot be determined. While both series may suggest cyclical behaviour, the trend would need to be removed to confirm this.

```{r page-1b TFR 1st diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(TFR$`Total Fertility Rate (TFR)`),
     type = 'l',
     main = "Plot of Differenced Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(TFR$`Total Fertility Rate (TFR)`)) # still not stationary
BirthAndFertility |>
  ACF(diff(TFR$`Total Fertility Rate (TFR)`), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(TFR$`Total Fertility Rate (TFR)`), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Fertility Rate (TFR)")
TFRm1610 <- Arima(TFR$`Total Fertility Rate (TFR)`, order=c(16,1,0))
TFRm1610 # AIC=-63.99 AICc=-45.99
res1610 <- residuals(TFRm1610)
acf(res1610, lag.max = 50, main = "ACF of Residuals for ARIMA(16,1,0) for TFR")
pacf(res1610, lag.max = 50, main = "PACF of Residuals for ARIMA(16,1,0) for TFR")
```
By using differencing, the plots look like the trend has been removed.

\newpage
```{r page-1c TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1611 <- Arima(TFR$`Total Fertility Rate (TFR)`, order=c(16,1,1))
TFRm1611 # AIC=-62.39 AICc=-41.66
res1611 <- residuals(TFRm1511)
acf(res1611, lag.max = 50, main = "ACF of Residuals for ARIMA(16,1,1) for TFR")
pacf(res1611, lag.max = 50, main = "PACF of Residuals for ARIMA(16,1,1) for TFR")
```

\newpage
```{r page-1d TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1511 <- Arima(TFR$`Total Fertility Rate (TFR)`, order=c(15,1,1))
TFRm1511 # AIC=-60.26 AICc=-42.26
res1511 <- residuals(TFRm1511)
acf(res1511, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,1) for TFR")
pacf(res1511, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,1) for TFR")
```

\newpage
```{r page-1e TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1413 <- Arima(TFR$`Total Fertility Rate (TFR)`, order=c(14,1,3))
TFRm1413 # AIC=-60.99 AICc=-40.26
res1413 <- residuals(TFRm1413)
acf(res1413, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,3) for TFR")
pacf(res1413, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,3) for TFR")
```




\newpage
```{r page-1e TFR 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(diff(TFR$`Total Fertility Rate (TFR)`)),
     type = 'l',
     main = "Plot of Differenced Twice Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(diff(TFR$`Total Fertility Rate (TFR)`))) # now stationary
BirthAndFertility |>
  ACF(diff(diff(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(diff(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Fertility Rate (TFR)")
```
\newpage
```{r page-2 TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm0 <- Arima(TFR$`Total Fertility Rate (TFR)`, order = c(16,1,0))
TFRm0 # to see the AIC
```
The ACF for TFR shows significant positive autocorrelations from lags 1 to 9, while TLB shows significant positive autocorrelations from lags 1 to 6, with a slow decrease in the ACF as the lag increases. This suggests that neither series is white noise, as there are multiple spikes outside the significance bounds.

The presence of a strong trend and slow decrease confirms that both TFR and TLB are non-stationary series. The absence of a sharp cut-off suggests that differencing or an autoregressive (AR) model may be suitable for this series.
An extended ACF plot for TFR confirms that significant positive autocorrelations persist up to lag 10.

ARIMA (16,1,0)

\newpage
```{r page-3 TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}

```
\newpage
```{r page-1 TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility,`Total Live-Births`)+
  labs(title = "Initial Plot of Total Live-Births")
```


```{r page-1b TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(BirthAndFertility$`Total Live-Births`), type = 'l') +
  labs(title = "Plot of Total Live Births (TLB)")
```

\newpage
```{r page-2 TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
BirthAndFertility |>
  ACF(diff(`Total Live-Births`), lag_max = 40) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Live-Births (TLB)")
BirthAndFertility |>
  PACF(diff(`Total Live-Births`), lag_max = 40) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Live-Births (TLB)")
```

ARIMA (13,1,0)

\newpage
```{r page-3 TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBm0 <- BirthAndFertility |>
  model(ARIMA(`Total Live-Births` ~ pdq(13,1,0)))
report(TLBm0) # to see the AIC
```
# Results


# Discussion


# Conclusion

# References

# Appendices

\newpage
# Research Questions
Can the ARIMA model accurately forecast the Singapore's Total Live Births (TLB) and Total Fertility Rate (TFR) from 2013 to 2024 given the data from 1960 to 2012, capturing the short-term fluctuations? 

# Statistical Appendix
The following statistical methods were applied throughout the analysis in the order presented below.

## Autoregressive Model (AR)
$$
y_t = c + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \cdots + \phi_p y_{t-p} + \varepsilon_t
$$
where:

\( y_t \) is the value at time t

\( \phi_i \) are model parameters

\( \varepsilon_t \) is white noise

Models the current values using its previous values (lags)

## Residuals
$$
e_t = y_t - \hat{y}_t
$$
\( y_t \) is the actual observed value at time t

\( \hat{y}_t \) is the predicted value at time t

The error which is the difference between the actual observed value and predicted value at time t, used to check the chosen model accounts for all the patterns

## Autocorrelation Function (ACF)
$$
\rho_k = \frac{\text{Cov}(y_t, y_{t-k})}{\text{Var}(y_t)}
$$
where:

\( \rho_k \) is the autocorrelation at lag \( k \)

Measures the correlation between a series \( y_t \) and lagged values \( y_{t-k} \), which checks for trend, seasonality and whether residuals are white noise

## Box-Cox Transformation
$$
y^{(\lambda)} =
\begin{cases}
\frac{y^\lambda - 1}{\lambda}, & \lambda \ne 0 \\
\log(y), & \lambda = 0
\end{cases}
$$

\(\lambda\) was estimated using the Guerrero method

## STL Decomposition
$$
y_t = T_t + R_t
$$
where:

\( T_t \) is the trend component,

\( R_t \) is the remainder

Decomposes time series into trend and remainder components, usually with the seasonality component as well but the data in this analysis is yearly

## Moving Average (MA)
$$
MA_k = \frac{1}{k} \sum_{i=-(k-1)/2}^{(k-1)/2} y_{t+i}
$$
k is the window size

In this analysis when k = 5, the moving average at time t is calculated using the 2 observations before and 2 observations after t. This process smooths out short-term fluctuations in the series

<https://github.com/twalepear/TSA.git>
