---
title: "Final Report"
author: "Thoa Thanh Thi Le"
output: pdf_document
---

# 1. Abstract
```{r echo = FALSE}
# 2/30 points
# comprehensively and succinctly summarised research question, analysis, and key findings.

# Statistical results translated into accessible and accurate description for non-scientific audience
```

# 2. Introduction
```{r echo = FALSE}
# 4/30 points Introduction and Literature Review
# clear, comprehensive, and highly relevant background information provided for data context

# research question clearly and succinctly stated, well-motivated by EDA, strongly linked to data context (e.g. using previous studies)
```
Total fertility rates (TFRs) are key population indicators that are widely monitored because they influence age structure and population growth (Wilmoth et al.2022; World Health Organisation). Globally, fertility rates have declined over recent decades (Chen et al. 2018; Shair et al. 2024; Vanella 2024). In Singapore, the total fertility rate has dropped below the replacement level required for a population to replace itself across generations (Jones & Hamid 2015; Tan, 2024; Wilmoth et al. 2022). Therefore, forecasting TFR and total live births (TLB) may provide insight into future fertility patterns and population change in Singapore. Preliminary exploratory data analysis indicated a long-term decline in TFR and TLB with periodic fluctuations, motivating the use of forecasting models capable of capturing these patterns.

Data from 1960 to 2024 on TLB and TFR were obtained from the Singapore Department of Statistics (2023). The research question addressed in this report is: Can ARIMA models accurately forecast Singapore's TLB and TFR while capturing short-term fluctuations?

This report is structured as follows. Part 3 describes the fertility decline in Singapore, Part 4 describes the process and workflow of obtaining viable ARIMA models, 

# 3. Literature Review

Around 1965, Singapore introduced policies aimed at reducing population growth, which contributed to subsequent declines in fertility rates (Chen et al. 2018; Goodkind 1996; Singh et al. 1988; Vanella 2024). Despite this long-term decline, there may be cyclical patterns in TFR and TLB. These fluctuations may be associated with cultural preferences surrounding Lunar zodiac Dragon years (1964, 1976, 1988, 2000, 2012 and 2024), which are considered especially auspicious years and have been associated with higher fertility rates and live births (Goodkind 1996; Jones & Hamid 2015). Forecasting such series therefore requires models that can capture both long-term trends and short-term variation. Among the various models for time series forecasting, ARIMA models are particularly flexible because they allow selection from a broad class of models rather than imposing a single model form on every time series (Kolambe & Arora 2024; Newbold 1983). Consequently, ARIMA models were selected as the forecasting approach for Singapore's TFR and TLB.

\newpage
# 4. Methods
```{r echo = FALSE}
# 5/30 points
# comprehensively described and justified modelling process and workflow in succinct manner

# model(s) used are highly appropriate for research question and data with comprehensive justification
```
For both TFR and TLB, an identical modelling workflow was applied. The analysis focused on the period from 1960-2012 for model fitting, with observations 2013-2024 reserved for evaluating forecast accuracy. Initial plots were used to examine the presence of trends, seasonality, and other notable patterns, which indicated a long-term decreasing trend in both series. Several transformations were considered including logarithmic transformation and first- and second-order differencing. Stationarity was assessed using the KPSS test. For each transformed series, autocorrelation (ACF) and partial autocorrelation (PACF) plots were used to find candidate ARIMA models. The candidate ARIMA models were then fitted and the residual ACF and PACF were examined to assess whether the residuals resembled white noise. Forecast accuracy was evaluated using mean square error (MSE) and mean absolute error (MAE) while model selection was guided by Akaike's Information Criterion (AIC) and the corrected Akaike's Information Criterion (AICc). This procedure was applied consistently to both series to ensure comparability of results.

\newpage
# 5. Results
```{r echo = FALSE}
# 5/30 points
# flawlessly estimated relevant time series models(s) and/or forecasts in R

# effectively and appropriately used composition of time series methods and advanced modelling for data

# critically interpreted and evaluated model fits with appropriate comprehensive use of diagnostics

```
\begin{center}
\textbf{Table 1. TFR - no log transformation}
\end{center}
| Model | AIC  | AICc | MSE | MAE |
|:--------:|:------:|:------:|:------:|:------:| 
| ARIMA(16,1,0) | -63.99 | -45.99 | 0.04753075 | 0.1690442 |**
| ARIMA(12,2,3) | -62.57 | -46.57 | 0.01218546 | 0.08274547 |**

\begin{center}
\textbf{Table 2. TFR - log transformed}
\end{center}
| Model | AIC  | AICc | MSE | MAE |
|:--------:|:------:|:------:|:------:|:------:| 
| ARIMA(13,1,3) | -136 | -118 | 0.02602064 | 0.121592 | **
| ARIMA(13,2,2) | -132.39 | -116.39 | 0.007426324 | 0.06826933 |**

\begin{center}
\textbf{Table 3. TLB - no log transformation}
\end{center}
| Model | AIC  | AICc | MSE | MAE |
|:--------:|:------:|:------:|:------:|:------:| 
| ARIMA(13,1,1) | 974.57 | 987.9 | 82856991 | 7698.072 |**
| ARIMA(13,1,2) | 973.01 | 988.56 | 54382525 | 6014.802 |**

\begin{center}
\textbf{Table 4. TLB - log transformed}
\end{center}
| Log Transformed TLB Model | AIC  | AICc | MSE | MAE |
|:--------:|:------:|:------:|:------:|:------:| 
| ARIMA(15,1,0) | -145.56 | -130.02 | 2239721240 | 47245.34 |**
| ARIMA(12,1,3) | -139.69 | -124.15 | 1752591080 | 41851.92 |**

# 6. Discussion
```{r echo = FALSE}
# 5/30 points
# thoroughly integrated limitations of models and processes into statistical conclusions

# comprehensively evaluated impact of model limitations on answering research question
```
For TFR, ARIMA(16,1,0) has the lowest AIC (-63.99) as the initial ARIMA model but higher MSE and MAE compared to the other models. However, the KPSS test shown that the data needed to be differenced twice. ARIMA(12,2,3) has the lowest AICc (-46.57) and relatively lower MSE and MAE which shows that the data were best to be differenced.

For the log transformed data, ARIMA(13,1,3) has the lowest AIC (-136) but the MSE and MAE were not as low as ARIMA(13,2,2) which has the lowest AICc (-116.39) and lowest MSE (0.007426) and lowest MAE (0.06827) among all TFR, non-log transformed and log transformed models overall. This indicated the data required to be log transformed and differenced.


# Conclusion
```{r echo = FALSE}
# 
```

```{r echo = FALSE}
# report presentation 2/30 points
# report structured with appropriate length, use of sections and highly professional style

# written with precise, concise, and correct scientific language with flawless grammar and professional tone

# visualisation presented with professional quality
```

\newpage
# References
```{r echo = FALSE}
# 5/30 points
# extension relevant literature reference, all correctly cited and referenced to in the main report

# appendix well presented as a coherent follow up to the main report
```

```{r echo = FALSE}
# (Chen et al. 2018; Shair et al. 2024; Vanella 2024)
```

```{r echo = FALSE}
# (Chen et al. 2018)
```
Chen, M., Yip, P.S.F. & Yap, M.T. (2018). Identifying the most Influential Groups in Determining Singapore’s Fertility. *Journal of Social Policy*, 47(1), 139-160. doi:10.1017/S0047279417000241

Goodkind, D. (1996). Chinese Lunar Birth Timing in Singapore: New Concerns for Child Quality amidst Multicultural Modernity. *Journal of Marriage and Family*, 58(3), 784-795. https://doi.org/10.2307/353736

Jones, G.W. & Hamid, W. (2015). Singapore's Pro-natalist Policies: To What Extend Have They Worked? *Low and Lower Fertility: Variations across Developed Countries*, 1, 33-61. Springer. https://doi.org/10.1007/978-3-319-21482-5_3

Kolambe, M. & Arora, S. (2024). Forecasting the Future: A Comprehensive Review of Time Series Prediction Techniques. *Journal of Electrical Systems*, 20(2), 575-586.

Newbold, P. (1983). ARIMA Model Building and the Time Series Analysis Approach to Forecasting. *Journal of Forecasting*, 2, 23-35. https://doi.org/10.1002/for.3980020104

Singapore Department of Statistics. (2023). *Births And Fertility Rates, Annual* [Dataset]. data.gov.sg. Accessed March 9, 2026, from https://data.gov.sg/datasets/d_e39eeaeadb571c0d0725ef1eec48d166/view

```{r echo = FALSE}
# (Shair et al. 2024)
```
Shair, S.N., Shaadan, N., Amirudin, N.A.B.M., & A’kashalf, N.A.M. (2024). Predictions of Malaysia Age-Specific Fertility Rates using the Lee-Carter and the Functional Data Approaches. *AIUB Journal of Science and Engineering*, 23(1), 71-78. doi: 10.53799/ajse.v23i1.1024

Singh, K., Viegas, O. & Ratnam, S.S. (1988). Fertility Trends in Singapore. *Journal of Biosocial Science*, 20(4), 401-409. https://doi.org/10.1017/S0021932000017521

Tan, J. (2024). Using Decomposed Total Fertility Rate (TFR) to Understand the Drivers for the Decline of TFR in Singapore. *Statistics Singapore Newsletter: Issue 1, 2024*, Household Income and Population Division, Singapore Department of Statistics. https://www.singstat.gov.sg/files/fc3d9271-e877-4a5c-a197-e244fddd80fb.pdf

```{r echo = FALSE}
# (Vanella 2024)
```
Vanella, P. (2016). The Total Fertility Rate in Germany until 2040 – A Stochastic Principal Components Projection based on Age-specific Fertility Rates. *Hannover Economic Papers
(HEP)*, 579. https://hdl.handle.net/10419/155634

```{r echo = FALSE}
# (Wilmoth et al. 2022)
```

Wilmoth, J., Menozzi, C. & Bassarsky, L. (2022). *Why population growth matters for sustainable development* (Policy Brief No. 130), United Nations Department of Economic and Social Affairs, Population Division.  https://www.un.org/development/desa/pd/sites/www.un.org.development.desa.pd/files/undesa_pd_2022_policy_brief_population_growth.pdf

World Health Organization. *Total fertility rate (per woman)*. Accessed June 7, 2026, from https://www.who.int/data/gho/indicator-metadata-registry/imr-details/123

\newpage
# Appendices
```{r page-0, echo = FALSE, warning = FALSE, message = FALSE}
library(fpp3) # time series analysis package
library(tsibble) # tsibble objects
library(lubridate) # for data to work with dates and times
library(feasts)
library(forecast) # for arima
library(latex2exp)
library(patchwork) # putting 2 plots next to each other

data <- readr::read_csv("BirthsAndFertilityRatesAnnual.csv") # read in the data
BirthAndFertility <- data |>
  pivot_longer( # change rows to column
    cols = `1960`:`2025`,
    names_to = "Year",
    values_to = "Value", # naming it to remember that it needs to be values
    values_transform = list(Value = as.numeric)
  ) |>
  filter(DataSeries %in% c("Total Fertility Rate (TFR)", "Total Live-Births")) |>
  mutate(Year = as.integer(Year)) |>
  filter(Year != 2025) |> # assignment specified until 2024
  pivot_wider( # taking TFR & Total Live-Births into 2 separate columns according to year
  names_from = DataSeries, 
  values_from = Value
  )
BirthAndFertility <- as_tsibble(BirthAndFertility, index = Year) # created a tsibble
# is_tsibble(BirthAndFertility) # checking it is a tsibble

TFR <- BirthAndFertility |>
  filter(Year <= 2012) |>
  select(Year, `Total Fertility Rate (TFR)`) |>
  as_tsibble(index = Year)
# is_tsibble(TFR)

TFR_ts <- ts(
  TFR$`Total Fertility Rate (TFR)`,
  start = min(TFR$Year),
  frequency = 1
)

actualTFR <- BirthAndFertility |>
  filter(Year %in% 2013:2024) |>
  pull(`Total Fertility Rate (TFR)`)

TLB <- BirthAndFertility |>
  filter(Year <= 2012) |>
  select(Year, `Total Live-Births`) |>
  as_tsibble(index = Year)
# is_tsibble(TLB)

TLB_ts <- ts(
  TLB$`Total Live-Births`,
  start = min(TLB$Year),
  frequency = 1
)

actualTLB <- BirthAndFertility |>
  filter(Year %in% 2013:2024) |>
  pull(`Total Live-Births`)
```

\newpage
```{r page-1 TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility, `Total Fertility Rate (TFR)`) +
  geom_vline(xintercept = 2012, colour = "red") +
  labs(title = "Figure 1. Initial Plot of Total Fertility Rate (TFR)")
```

```{r page-1b TFR 1st diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(TFR$`Total Fertility Rate (TFR)`),
     type = 'l',
     main = "Plot of Differenced Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(TFR$`Total Fertility Rate (TFR)`)) # still not stationary
BirthAndFertility |>
  ACF(diff(TFR$`Total Fertility Rate (TFR)`), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(TFR$`Total Fertility Rate (TFR)`), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Fertility Rate (TFR)")
TFRm1610 <- Arima(TFR_ts, order=c(16,1,0))
TFRm1610 # AIC=-63.99 AICc=-45.99
res1610 <- residuals(TFRm1610)
acf(res1610, lag.max = 50, main = "ACF of Residuals for ARIMA(16,1,0) for TFR")
pacf(res1610, lag.max = 50, main = "PACF of Residuals for ARIMA(16,1,0) for TFR")
arima1610 <- forecast(TFRm1610, h = 12)
pred1610 <- arima1610$mean
mse1610 <- mean((actualTFR - pred1610)^2) # 0.04753075
abs1610 <- mean(abs(actualTFR - pred1610)) # 0.1690442
```

\newpage
```{r page-1c TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRm1611 <- Arima(TFR_ts, order=c(16,1,1))
# TFRm1611 # AIC=-62.39 AICc=-41.66
# res1611 <- residuals(TFRm1611)
# acf(res1611, lag.max = 50, main = "ACF of Residuals for ARIMA(16,1,1) for TFR")
# pacf(res1611, lag.max = 50, main = "PACF of Residuals for ARIMA(16,1,1) for TFR")
# arima1611 <- forecast(TFRm1611, h = 12)
# pred1611 <- arima1611$mean
# mse1611 <- mean((actualTFR - pred1611)^2) # 0.04783471
# abs1611 <- mean(abs(actualTFR - pred1611)) # 0.1700046
```

\newpage
```{r page-1d TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRm1511 <- Arima(TFR_ts, order=c(15,1,1))
# TFRm1511 # AIC=-60.26 AICc=-42.26
# res1511 <- residuals(TFRm1511)
# acf(res1511, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,1) for TFR")
# pacf(res1511, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,1) for TFR")
# arima1511 <- forecast(TFRm1511, h = 12)
# pred1511 <- arima1511$mean
# mse1511 <- mean((actualTFR - pred1511)^2) # 0.08431368
# abs1511 <- mean(abs(actualTFR - pred1511)) # 0.223524
```

\newpage
```{r page-1e TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRm1413 <- Arima(TFR_ts, order=c(14,1,3))
# TFRm1413 # AIC=-60.99 AICc=-40.26
# res1413 <- residuals(TFRm1413)
# acf(res1413, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,3) for TFR")
# pacf(res1413, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,3) for TFR")
# arima1413 <- forecast(TFRm1413, h = 12)
# pred1413 <- arima1413$mean
# mse1413 <- mean((actualTFR - pred1413)^2) # 0.01899351
# abs1413 <- mean(abs(actualTFR - pred1413)) # 0.09964934
```

\newpage
```{r page-1f TFR 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(diff(TFR$`Total Fertility Rate (TFR)`)),
     type = 'l',
     main = "Plot of Differenced Twice Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(diff(TFR$`Total Fertility Rate (TFR)`))) # now stationary
BirthAndFertility |>
  ACF(diff(diff(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Twice Differenced Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(diff(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Twice Differenced Total Fertility Rate (TFR)")
TFRm1220 <- Arima(TFR_ts, order=c(12,2,0)) # not a viable model
TFRm1220 # AIC=-57.55 AICc=-47.72
res1220 <- residuals(TFRm1220)
acf(res1220, lag.max = 50, main = "ACF of Residuals for ARIMA(12,2,0) for TFR")
pacf(res1220, lag.max = 50, main = "PACF of Residuals for ARIMA(12,2,0) for TFR")
```

\newpage
```{r page-1g TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1223 <- Arima(TFR_ts, order=c(12,2,3))
TFRm1223 # AIC=-62.57 AICc=-46.57
res1223 <- residuals(TFRm1223)
acf(res1223, lag.max = 50, main = "ACF of Residuals for ARIMA(12,2,3) for TFR")
pacf(res1223, lag.max = 50, main = "PACF of Residuals for ARIMA(12,2,3) for TFR")
arima1223 <- forecast(TFRm1223, h = 12)
pred1223 <- arima1223$mean
mse1223 <- mean((actualTFR - pred1223)^2) # 0.01218546
abs1223 <- mean(abs(actualTFR - pred1223)) # 0.08274547
```

\newpage
```{r page-1h TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRm1123 <- Arima(TFR_ts, order=c(11,2,3))
# TFRm1123 # AIC=-52.91 AICc=-39.19
# res1123 <- residuals(TFRm1123)
# acf(res1123, lag.max = 50, main = "ACF of Residuals for ARIMA(11,2,3) for TFR")
# pacf(res1123, lag.max = 50, main = "PACF of Residuals for ARIMA(11,2,3) for TFR")
# arima1123 <- forecast(TFRm1123, h = 12)
# pred1123 <- arima1123$mean
# mse1123 <- mean((actualTFR - pred1123)^2) # 0.05260825
# abs1123 <- mean(abs(actualTFR - pred1123)) # 0.1823268
```

\newpage
```{r page-1i TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRm1023 <- Arima(TFR_ts, order=c(10,2,3))
# TFRm1023 # AIC=-54.69 AICc=-43.02
# res1023 <- residuals(TFRm1023)
# acf(res1023, lag.max = 50, main = "ACF of Residuals for ARIMA(10,2,3) for TFR")
# pacf(res1023, lag.max = 50, main = "PACF of Residuals for ARIMA(10,2,3) for TFR")
# arima1023 <- forecast(TFRm1023, h = 12)
# pred1023 <- arima1023$mean
# mse1023 <- mean((actualTFR - pred1023)^2) # 0.05680021
# abs1023 <- mean(abs(actualTFR - pred1023)) # 0.1898403
```

\newpage
```{r page-1j TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(log(TFR$`Total Fertility Rate (TFR)`),
     type = 'l',
     main = "Plot of Log Transformation for Total Fertility Rate (TFR) until 2012")
plot(diff(log(TFR$`Total Fertility Rate (TFR)`)),
     type = 'l',
     main = "Plot of Differenced Log Transformation for Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(log(TFR$`Total Fertility Rate (TFR)`))) # borderline 0.0405
BirthAndFertility |>
  ACF(diff(log(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Log Transformation of Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(log(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Log Transformation of Total Fertility Rate (TFR)")
TFRml1310 <- Arima(log(TFR_ts), order=c(13,1,0)) # not a viable model
TFRml1310 # AIC=-131.8 AICc=-120.45
resl1310 <- residuals(TFRml1310)
acf(resl1310, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,0) for TFR with log transformation")
pacf(resl1310, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,0) for TFR with log transformation")
```

\newpage
```{r page-1k TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1313 <- Arima(log(TFR_ts), order=c(13,1,3))
TFRml1313 # AIC=-136 AICc=-118
resl1313 <- residuals(TFRml1313)
acf(resl1313, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,3) for TFR with log transformation")
pacf(resl1313, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,3) for TFR with log transformation")
arimal1313 <- forecast(TFRml1313, h = 12)
predl1313 <- arimal1313$mean
msel1313 <- mean((actualTFR - exp(predl1313))^2) # 0.02602064
absl1313 <- mean(abs(actualTFR - exp(predl1313))) # 0.121592
```

\newpage
```{r page-1l TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRml1214 <- Arima(log(TFR_ts), order=c(12,1,4))
# TFRml1214 # AIC=-131.83 AICc=-113.83
# resl1214 <- residuals(TFRml1214)
# acf(resl1214, lag.max = 50, main = "ACF of Residuals for ARIMA(12,1,4) for TFR with log transformation")
# pacf(resl1214, lag.max = 50, main = "PACF of Residuals for ARIMA(12,1,4) for TFR with log transformation")
# arimal1214 <- forecast(TFRml1214, h = 12)
# predl1214 <- arimal1214$mean
# msel1214 <- mean((actualTFR - exp(predl1214))^2) # 0.02101819
# absl1214 <- mean(abs(actualTFR - exp(predl1214))) # 0.1132223
```

\newpage
```{r page-1m TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRml1315 <- Arima(log(TFR_ts), order=c(13,1,5))
# TFRml1315 # AIC=-133.96 AICc=-110.21
# resl1315 <- residuals(TFRml1315)
# acf(resl1315, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,5) for TFR with log transformation")
# pacf(resl1315, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,5) for TFR with log transformation")
# arimal1315 <- forecast(TFRml1315, h = 12)
# predl1315 <- arimal1315$mean
# msel1315 <- mean((actualTFR - exp(predl1315))^2) # 0.05826379
# absl1315 <- mean(abs(actualTFR - exp(predl1315))) # 0.1840232
```

\newpage
```{r page-1n TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRml1215 <- Arima(log(TFR_ts), order=c(12,1,5))
# TFRml1215 # AIC=-130.02 AICc=-109.3
# resl1215 <- residuals(TFRml1215)
# acf(resl1215, lag.max = 50, main = "ACF of Residuals for ARIMA(12,1,5) for TFR with log transformation")
# pacf(resl1215, lag.max = 50, main = "PACF of Residuals for ARIMA(12,1,5) for TFR with log transformation")
# arimal1215 <- forecast(TFRml1215, h = 12)
# predl1215 <- arimal1215$mean
# msel1215 <- mean((actualTFR - exp(predl1215))^2) # 0.02172126
# absl1215 <- mean(abs(actualTFR - exp(predl1215))) # 0.1147922
```

\newpage
```{r page-1o TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))),
     type = 'l',
     main = "Plot of Differenced Twice Log Transformation for Total Fertility Rate (TFR)")
unitroot_kpss(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))))
BirthAndFertility |>
  ACF(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Twice Differenced Log Transformation of Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Twice Differenced Log Transformation of Total Fertility Rate (TFR)")
TFRml1520 <- Arima(log(TFR_ts), order=c(15,2,0))
TFRml1520 # AIC=-133.67 AICc=-117.67
resl1520 <- residuals(TFRml1520)
acf(resl1520, lag.max = 50, main = "ACF of Residuals for ARIMA(15,2,0) for TFR with log transformation")
pacf(resl1520, lag.max = 50, main = "PACF of Residuals for ARIMA(15,2,0) for TFR with log transformation")
arimal1520 <- forecast(TFRml1520, h = 12)
predl1520 <- arimal1520$mean
msel1520 <- mean((actualTFR - exp(predl1520))^2) # 0.01869052
absl1520 <- mean(abs(actualTFR - exp(predl1520))) # 0.1056717
```

\newpage
```{r page-1p TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRml1522 <- Arima(log(TFR_ts), order=c(15,2,2))
# TFRml1522 # AIC=-135.54 AICc=-114.17
# resl1522 <- residuals(TFRml1522)
# acf(resl1522, lag.max = 50, main = "ACF of Residuals for ARIMA(15,2,2) for TFR with log transformation")
# pacf(resl1522, lag.max = 50, main = "PACF of Residuals for ARIMA(15,2,2) for TFR with log transformation")
# arimal1522 <- forecast(TFRml1522, h = 12)
# predl1522 <- arimal1522$mean
# msel1522 <- mean((actualTFR - exp(predl1522))^2) # 0.02235162
# absl1522 <- mean(abs(actualTFR - exp(predl1522))) # 0.1187815
```

\newpage
```{r page-1q TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRml1422 <- Arima(log(TFR_ts), order=c(14,2,2))
# TFRml1422 # AIC=-130.49 AICc=-111.94
# resl1422 <- residuals(TFRml1422)
# acf(resl1422, lag.max = 50, main = "ACF of Residuals for ARIMA(14,2,2) for TFR with log transformation")
# pacf(resl1422, lag.max = 50, main = "PACF of Residuals for ARIMA(14,2,2) for TFR with log transformation")
# arimal1422 <- forecast(TFRml1422, h = 12)
# predl1422 <- arimal1422$mean
# msel1422 <- mean((actualTFR - exp(predl1422))^2) # 0.01087075
# absl1422 <- mean(abs(actualTFR - exp(predl1422))) # 0.08352761
```

\newpage
```{r page-1r TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1322 <- Arima(log(TFR_ts), order=c(13,2,2))
TFRml1322 # AIC=-132.39 AICc=-116.39
resl1322 <- residuals(TFRml1322)
acf(resl1322, lag.max = 50, main = "ACF of Residuals for ARIMA(13,2,2) for TFR with log transformation")
pacf(resl1322, lag.max = 50, main = "PACF of Residuals for ARIMA(13,2,2) for TFR with log transformation")
arimal1322 <- forecast(TFRml1322, h = 12)
predl1322 <- arimal1322$mean
msel1322 <- mean((actualTFR - exp(predl1322))^2) # 0.007426324
absl1322 <- mean(abs(actualTFR - exp(predl1322))) # 0.06826933
```

\newpage
```{r page-1s TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TFRml1222 <- Arima(log(TFR_ts), order=c(12,2,2))
# TFRml1222 # AIC=-127.34 AICc=-113.62
# resl1222 <- residuals(TFRml1222)
# acf(resl1222, lag.max = 50, main = "ACF of Residuals for ARIMA(12,2,2) for TFR with log transformation")
# pacf(resl1222, lag.max = 50, main = "PACF of Residuals for ARIMA(12,2,2) for TFR with log transformation")
# arimal1222 <- forecast(TFRml1222, h = 12)
# predl1222 <- arimal1222$mean
# msel1222 <- mean((actualTFR - exp(predl1222))^2) # 0.0498791
# absl1222 <- mean(abs(actualTFR - exp(predl1222))) # 0.1720094
```

\newpage
```{r page-2 TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility,`Total Live-Births`) +
  geom_vline(xintercept = 2012, colour = "red") +
  labs(title = "Initial Plot of Total Live-Births")
```

```{r page-2b TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(TLB$`Total Live-Births`),
     type = 'l',
     main = "Plot of Differenced Total Live Births (TLB) until 2012")
unitroot_kpss(diff(TLB$`Total Live-Births`)) # stationary
BirthAndFertility |>
  ACF(diff(TLB$`Total Live-Births`), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Live Births (TLB)")
BirthAndFertility |>
  PACF(diff(TLB$`Total Live-Births`), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Live Births (TLB)")
TLBm1410 <- Arima(TLB_ts, order=c(14,1,0))
TLBm1410 # AIC=975.33 AICc=988.66
resb1410 <- residuals(TLBm1410)
acf(resb1410, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,0) for TLB")
pacf(resb1410, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,0) for TLB")
arima1410 <- forecast(TLBm1410, h = 12)
pred1410 <- arima1410$mean
mse1410 <- mean((actualTLB - pred1410)^2) # 67367429
abs1410 <- mean(abs(actualTLB - pred1410)) # 6853.008
```

\newpage
```{r page-2c TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TLBm1411 <- Arima(TLB_ts, order=c(14,1,1))
# TLBm1411 # AIC=975.62 AICc=991.17
# resb1411 <- residuals(TLBm1411)
# acf(resb1411, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,1) for TLB")
# pacf(resb1411, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,1) for TLB")
# arima1411 <- forecast(TLBm1411, h = 12)
# pred1411 <- arima1411$mean
# mse1411 <- mean((actualTLB - pred1411)^2) # 84179161
# abs1411 <- mean(abs(actualTLB - pred1411)) # 7769.997
```

\newpage
```{r page-2d TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBm1311 <- Arima(TLB_ts, order=c(13,1,1))
TLBm1311 # AIC=974.57 AICc=987.9
resb1311 <- residuals(TLBm1311)
acf(resb1311, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,1) for TLB")
pacf(resb1311, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,1) for TLB")
arima1311 <- forecast(TLBm1311, h = 12)
pred1311 <- arima1311$mean
mse1311 <- mean((actualTLB - pred1311)^2) # 82856991
abs1311 <- mean(abs(actualTLB - pred1311)) # 7698.072
```

\newpage
```{r page-2e TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBm1312 <- Arima(TLB_ts, order=c(13,1,2))
TLBm1312 # AIC=973.01 AICc=988.56
resb1312 <- residuals(TLBm1312)
acf(resb1312, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,2) for TLB")
pacf(resb1312, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,2) for TLB")
arima1312 <- forecast(TLBm1312, h = 12)
pred1312 <- arima1312$mean
mse1312 <- mean((actualTLB - pred1312)^2) # 54382525
abs1312 <- mean(abs(actualTLB - pred1312)) # 6014.802
```

\newpage
```{r page-2f TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(log(TLB$`Total Live-Births`),
     type = 'l',
     main = "Plot of Log Transformation for Total Live-Births (TLB) until 2012")
plot(diff(log(TLB$`Total Live-Births`)),
     type = 'l',
     main = "Plot of Differenced Log Transformation for Total Live-Births (TLB) until 2012")
unitroot_kpss(diff(log(TLB$`Total Live-Births`))) # stationary
BirthAndFertility |>
  ACF(diff(log(TLB$`Total Live-Births`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Log Transformation of Total Live-Births (TLB)")
BirthAndFertility |>
  PACF(diff(log(TLB$`Total Live-Births`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Log Transformation of Total Live-Births (TLB)")
TLBml1510 <- Arima(log(TLB_ts), order=c(15,1,0))
TLBml1510 # AIC=-145.56 AICc=-130.02
resl1510 <- residuals(TLBml1510)
acf(resl1510, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,0) for TLB with log transformation")
pacf(resl1510, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,0) for TLB with log transformation")
arimal1510 <- forecast(TLBml1510, h = 12)
predl1510 <- arimal1510$mean
msel1510 <- mean((actualTFR - exp(predl1510))^2) # 2239721240
absl1510 <- mean(abs(actualTFR - exp(predl1510))) # 47245.34
```
\newpage
```{r page-2g TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TLBml1511 <- Arima(log(TLB_ts), order=c(15,1,1))
# TLBml1511 # AIC=-145.43 AICc=-127.43
# resl1511 <- residuals(TLBml1511)
# acf(resl1511, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,1) for TLB with log transformation")
# pacf(resl1511, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,1) for TLB with log transformation")
# arimal1511 <- forecast(TLBml1511, h = 12)
# predl1511 <- arimal1511$mean
# msel1511 <- mean((actualTFR - exp(predl1511))^2) # 2173041601
# absl1511 <- mean(abs(actualTFR - exp(predl1511))) # 46542.86
```

\newpage
```{r page-2h TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TLBml1411 <- Arima(log(TLB_ts), order=c(14,1,1))
# TLBml1411 # AIC=-140.56 AICc=-125.02
# resl1411 <- residuals(TLBml1411)
# acf(resl1411, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,1) for TLB with log transformation")
# pacf(resl1411, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,1) for TLB with log transformation")
# arimal1411 <- forecast(TLBml1411, h = 12)
# predl1411 <- arimal1411$mean
# msel1411 <- mean((actualTFR - exp(predl1411))^2) # 2173041601
# absl1411 <- mean(abs(actualTFR - exp(predl1411))) # 46542.86
```

\newpage
```{r page-2i TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
# TLBml1311 <- Arima(log(TLB_ts), order=c(13,1,1))
# TLBml1311 # AIC=-141.45 AICc=-128.11
# resl1311 <- residuals(TLBml1311)
# acf(resl1311, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,1) for TLB with log transformation")
# pacf(resl1311, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,1) for TLB with log transformation")
# arimal1311 <- forecast(TLBml1311, h = 12)
# predl1311 <- arimal1311$mean
# msel1311 <- mean((actualTFR - exp(predl1311))^2) # 2200303964
# absl1311 <- mean(abs(actualTFR - exp(predl1311))) # 46836.59
```

\newpage
```{r page-2j TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBml1213 <- Arima(log(TLB_ts), order=c(12,1,3))
TLBml1213 # AIC=-139.69 AICc=-124.15
resl1213 <- residuals(TLBml1213)
acf(resl1213, lag.max = 50, main = "ACF of Residuals for ARIMA(12,1,3) for TLB with log transformation")
pacf(resl1213, lag.max = 50, main = "PACF of Residuals for ARIMA(12,1,3) for TLB with log transformation")
arimal1213 <- forecast(TLBml1213, h = 12)
predl1213 <- arimal1213$mean
msel1213 <- mean((actualTFR - exp(predl1213))^2) # 1752591080
absl1213 <- mean(abs(actualTFR - exp(predl1213))) # 41851.92
```
## Autoregressive Model Integrated (ARIMA) model
$$
\phi(B)(1-B)^d y_t = c + \theta(B)\varepsilon_t
$$
\newpage
\begin{center}
\textbf{Table 5. Full table for TFR - no log transformation}
\end{center}
| Model | AIC  | AICc | MSE | MAE |   |
|:--------:|:------:|:------:|:------:|:------:|:------:|
| ARIMA(16,1,0) | -63.99 | -45.99 | 0.04753075 | 0.1690442 | Lowest AIC |
| ARIMA(16,1,1) | -62.39 | -42.66 | 0.04783471 | 0.1700046 |  |
| ARIMA(15,1,1) | -60.26 | -42.26 | 0.08431368 | 0.223524 |  |
| ARIMA(14,1,3) | -60.99 | -40.26 | 0.01899351 | 0.09964934 |  |
| ARIMA(12,2,3) | -62.57 | -46.57 | 0.01218546 | 0.08274547 | Lowest AICc |
| ARIMA(11,2,3) | -52.91 | -39.19 | 0.05260825 | 0.1823268 |
| ARIMA(10,2,3) | -54.69 | -43.02 | 0.05680021 | 0.1898403 |

\begin{center}
\textbf{Table 6. Full table for TFR - log transformed}
\end{center}
| Model | AIC  | AICc | MSE | MAE |   |
|:--------:|:------:|:------:|:------:|:------:|:------:|
| ARIMA(13,1,3) | -136 | -118 | 0.02602064 | 0.121592 | Lowest AIC Log Transform |
| ARIMA(12,1,4) | -131.83 | -113.83 | 0.02101819 | 0.1132223 |  |
| ARIMA(13,1,5) | -133.96 | -110.21 | 0.05826379 | 0.1840232 |  |
| ARIMA(12,1,5) | -130.02 | -109.3 | 0.02172126 | 0.1147922 |  |
| ARIMA(15,2,0) | -133.67 | -117.67 | 0.01869052 | 0.1056717 |  |
| ARIMA(15,2,2) | -135.54 | -114.17 | 0.02235162 | 0.1187815 |  |
| ARIMA(14,2,2) | -130.49 | -111.94 | 0.01087075 | 0.08352761 |  |
| ARIMA(13,2,2) | -132.39 | -116.39 | 0.007426324 | 0.06826933 | Lowest MSE & MAE |
| ARIMA(12,2,2) | -127.34 | -113.62 | 0.0498791 | 0.1720094 |

\begin{center}
\textbf{Table 7. Full table for TLB - no log transformation}
\end{center}
| Model | AIC  | AICc | MSE | MAE |  |
|:--------:|:------:|:------:|:------:|:------:|:------:|
| ARIMA(14,1,0) | 975.33 | 988.66 | 67367429 | 6853.008 |  |
| ARIMA(14,1,1) | 975.62 | 991.17 | 84179161 | 7769.997 |  |
| ARIMA(13,1,1) | 974.57 | 987.9 | 82856991 | 7698.072 | lowest AICc (no log transformation) |
| ARIMA(13,1,2) | 973.01 | 988.56 | 54382525 | 6014.802 | lowest AIC (no log transformation), MSE & MAE |

\newpage
\begin{center}
\textbf{Table 8. Full table for TLB - log transformed}
\end{center}
| Model | AIC  | AICc | MSE | MAE |  |
|:--------:|:------:|:------:|:------:|:------:|:------:|
| ARIMA(15,1,0) | -145.56 | -130.02 | 2239721240 | 47245.34 | lowest AICc (log transformed) |
| ARIMA(15,1,1) | -145.43 | -127.43 | 2173041601 | 46542.86 |  |
| ARIMA(14,1,1) | -140.56 | -125.02 | 2211919788 | 46958.53 |  |
| ARIMA(13,1,1) | -141.45 | -128.11 | 2200303964 | 46836.59 |  |
| ARIMA(12,1,3) | -139.69 | -124.15 | 1752591080 | 41851.92 |  |

<https://github.com/twalepear/TSA.git>
