---
title: "1.2 Final Report"
author: "Thoa Le (a1847933)"
output: pdf_document
---

# Abstract
```{r echo = FALSE}
# 2/30 points
# comprehensively and succinctly summarised research question, analysis, and key findings.

# Statistical results translated into accessible and accurate description for non-scientific audience
```


# Introduction and Literature Review
```{r echo = FALSE}
# 4/30 points
# clear, comprehensive, and highly relevant background information provided for data context

# research question clearly and succinctly stated, well-motivated by EDA, strongly linked to data context (e.g. using previous studies)
```

# Methods
```{r echo = FALSE}
# 5/30 points
# comprehensively described and justified modelling process and workflow in succinct manner

# model(s) used are highly appropriate for research question and data with comprehensive justification
```
Data from 1960 to 2024 on Total Live Births (TLB) and Total Fertility Rates (TFR) was pulled from Singapore Department of Statistics (data.gov.sg). The data was then transformed into a tsibble object with three columns: Year, Total Fertility Rates (TFR) and Total Live Births.

```{r page-0, echo = FALSE, warning = FALSE, message = FALSE}
library(fpp3) # time series analysis package
library(tsibble) # tsibble objects
library(lubridate) # for data to work with dates and times
library(feasts)
library(forecast) # for arima
library(latex2exp)
library(patchwork) # putting 2 plots next to each other

data <- readr::read_csv("BirthsAndFertilityRatesAnnual.csv") # read in the data
BirthAndFertility <- data |>
  pivot_longer( # change rows to column
    cols = `1960`:`2025`,
    names_to = "Year",
    values_to = "Value", # naming it to remember that it needs to be values
    values_transform = list(Value = as.numeric)
  ) |>
  filter(DataSeries %in% c("Total Fertility Rate (TFR)", "Total Live-Births")) |>
  mutate(Year = as.integer(Year)) |>
  filter(Year != 2025) |> # assignment specified until 2024
  pivot_wider( # taking TFR & Total Live-Births into 2 separate columns according to year
  names_from = DataSeries, 
  values_from = Value
  )
BirthAndFertility <- as_tsibble(BirthAndFertility, index = Year) # created a tsibble
# is_tsibble(BirthAndFertility) # checking it is a tsibble

TFR <- BirthAndFertility |>
  filter(Year <= 2012) |>
  select(Year, `Total Fertility Rate (TFR)`) |>
  as_tsibble(index = Year)
# is_tsibble(TFR)

TFR_ts <- ts(
  TFR$`Total Fertility Rate (TFR)`,
  start = min(TFR$Year),
  frequency = 1
)

actualTFR <- BirthAndFertility |>
  filter(Year %in% 2013:2024) |>
  pull(`Total Fertility Rate (TFR)`)

TLB <- BirthAndFertility |>
  filter(Year <= 2012) |>
  select(Year, `Total Live-Births`) |>
  as_tsibble(index = Year)
# is_tsibble(TLB)

TLB_ts <- ts(
  TLB$`Total Live-Births`,
  start = min(TLB$Year),
  frequency = 1
)

actualTLB <- BirthAndFertility |>
  filter(Year %in% 2013:2024) |>
  pull(`Total Live-Births`)
```

The TFR was initially plotted to examine the presence of trends, seasonality, and other notable patterns (Figure 1). The subsequent analysis focuses on the period from 1960 to 2012, a second plot was produced for visualisation and modelling purposes (Figure 2). 

There is a long-term decreasing trend over time hence the data was differenced. Viable ARIMA models were found but the KPSS unit root test suggested the data needed to be differenced again where more viable ARIMA models were found. The initial plot decreasing trend suggests a log transformation would be appropriate where more models were found after differencing and twice differencing.

\newpage
```{r page-1 TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility, `Total Fertility Rate (TFR)`) +
  labs(title = "Figure 1. Initial Plot of Total Fertility Rate (TFR)")
autoplot(TFR) +
  labs(title = "Figure 2. Initial Plot of Total Fertility Rate (TFR) until 2012")
```
Figure 1 show a long-term decreasing trend over time



```{r page-1b TFR 1st diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(TFR$`Total Fertility Rate (TFR)`),
     type = 'l',
     main = "Plot of Differenced Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(TFR$`Total Fertility Rate (TFR)`)) # still not stationary
BirthAndFertility |>
  ACF(diff(TFR$`Total Fertility Rate (TFR)`), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(TFR$`Total Fertility Rate (TFR)`), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Fertility Rate (TFR)")
TFRm1610 <- Arima(TFR_ts, order=c(16,1,0))
TFRm1610 # AIC=-63.99 AICc=-45.99
res1610 <- residuals(TFRm1610)
acf(res1610, lag.max = 50, main = "ACF of Residuals for ARIMA(16,1,0) for TFR")
pacf(res1610, lag.max = 50, main = "PACF of Residuals for ARIMA(16,1,0) for TFR")
arima1610 <- forecast(TFRm1610, h = 12)
pred1610 <- arima1610$mean
mse1610 <- mean((actualTFR - pred1610)^2) # 0.04753075
abs1610 <- mean(abs(actualTFR - pred1610)) # 0.1690442
```
By using differencing, the plots look like the trend has been removed.

\newpage
```{r page-1c TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1611 <- Arima(TFR_ts, order=c(16,1,1))
TFRm1611 # AIC=-62.39 AICc=-41.66
res1611 <- residuals(TFRm1611)
acf(res1611, lag.max = 50, main = "ACF of Residuals for ARIMA(16,1,1) for TFR")
pacf(res1611, lag.max = 50, main = "PACF of Residuals for ARIMA(16,1,1) for TFR")
arima1611 <- forecast(TFRm1611, h = 12)
pred1611 <- arima1611$mean
mse1611 <- mean((actualTFR - pred1611)^2) # 0.04783471
abs1611 <- mean(abs(actualTFR - pred1611)) # 0.1700046
```

\newpage
```{r page-1d TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1511 <- Arima(TFR_ts, order=c(15,1,1))
TFRm1511 # AIC=-60.26 AICc=-42.26
res1511 <- residuals(TFRm1511)
acf(res1511, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,1) for TFR")
pacf(res1511, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,1) for TFR")
arima1511 <- forecast(TFRm1511, h = 12)
pred1511 <- arima1511$mean
mse1511 <- mean((actualTFR - pred1511)^2) # 0.08431368
abs1511 <- mean(abs(actualTFR - pred1511)) # 0.223524
```

\newpage
```{r page-1e TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1413 <- Arima(TFR_ts, order=c(14,1,3))
TFRm1413 # AIC=-60.99 AICc=-40.26
res1413 <- residuals(TFRm1413)
acf(res1413, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,3) for TFR")
pacf(res1413, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,3) for TFR")
arima1413 <- forecast(TFRm1413, h = 12)
pred1413 <- arima1413$mean
mse1413 <- mean((actualTFR - pred1413)^2) # 0.01899351
abs1413 <- mean(abs(actualTFR - pred1413)) # 0.09964934
```

\newpage
```{r page-1f TFR 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(diff(TFR$`Total Fertility Rate (TFR)`)),
     type = 'l',
     main = "Plot of Differenced Twice Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(diff(TFR$`Total Fertility Rate (TFR)`))) # now stationary
BirthAndFertility |>
  ACF(diff(diff(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Twice Differenced Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(diff(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Twice Differenced Total Fertility Rate (TFR)")
TFRm1220 <- Arima(TFR_ts, order=c(12,2,0)) # not a viable model
TFRm1220 # AIC=-57.55 AICc=-47.72
res1220 <- residuals(TFRm1220)
acf(res1220, lag.max = 50, main = "ACF of Residuals for ARIMA(12,2,0) for TFR")
pacf(res1220, lag.max = 50, main = "PACF of Residuals for ARIMA(12,2,0) for TFR")
```

\newpage
```{r page-1g TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1223 <- Arima(TFR_ts, order=c(12,2,3))
TFRm1223 # AIC=-62.57 AICc=-46.57
res1223 <- residuals(TFRm1223)
acf(res1223, lag.max = 50, main = "ACF of Residuals for ARIMA(12,2,3) for TFR")
pacf(res1223, lag.max = 50, main = "PACF of Residuals for ARIMA(12,2,3) for TFR")
arima1223 <- forecast(TFRm1223, h = 12)
pred1223 <- arima1223$mean
mse1223 <- mean((actualTFR - pred1223)^2) # 0.01218546
abs1223 <- mean(abs(actualTFR - pred1223)) # 0.08274547
```

\newpage
```{r page-1h TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1123 <- Arima(TFR_ts, order=c(11,2,3))
TFRm1123 # AIC=-52.91 AICc=-39.19
res1123 <- residuals(TFRm1123)
acf(res1123, lag.max = 50, main = "ACF of Residuals for ARIMA(11,2,3) for TFR")
pacf(res1123, lag.max = 50, main = "PACF of Residuals for ARIMA(11,2,3) for TFR")
arima1123 <- forecast(TFRm1123, h = 12)
pred1123 <- arima1123$mean
mse1123 <- mean((actualTFR - pred1123)^2) # 0.05260825
abs1123 <- mean(abs(actualTFR - pred1123)) # 0.1823268
```

\newpage
```{r page-1i TFR, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRm1023 <- Arima(TFR_ts, order=c(10,2,3))
TFRm1023 # AIC=-54.69 AICc=-43.02
res1023 <- residuals(TFRm1023)
acf(res1023, lag.max = 50, main = "ACF of Residuals for ARIMA(10,2,3) for TFR")
pacf(res1023, lag.max = 50, main = "PACF of Residuals for ARIMA(10,2,3) for TFR")
arima1023 <- forecast(TFRm1023, h = 12)
pred1023 <- arima1023$mean
mse1023 <- mean((actualTFR - pred1023)^2) # 0.05680021
abs1023 <- mean(abs(actualTFR - pred1023)) # 0.1898403
```

\newpage
```{r page-1j TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(log(TFR$`Total Fertility Rate (TFR)`),
     type = 'l',
     main = "Plot of Log Transformation for Total Fertility Rate (TFR) until 2012")
plot(diff(log(TFR$`Total Fertility Rate (TFR)`)),
     type = 'l',
     main = "Plot of Differenced Log Transformation for Total Fertility Rate (TFR) until 2012")
unitroot_kpss(diff(log(TFR$`Total Fertility Rate (TFR)`))) # borderline 0.0405
BirthAndFertility |>
  ACF(diff(log(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Log Transformation of Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(log(TFR$`Total Fertility Rate (TFR)`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Log Transformation of Total Fertility Rate (TFR)")
TFRml1310 <- Arima(log(TFR_ts), order=c(13,1,0)) # not a viable model
TFRml1310 # AIC=-131.8 AICc=-120.45
resl1310 <- residuals(TFRml1310)
acf(resl1310, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,0) for TFR with log transformation")
pacf(resl1310, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,0) for TFR with log transformation")
```

\newpage
```{r page-1k TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1313 <- Arima(log(TFR_ts), order=c(13,1,3))
TFRml1313 # AIC=-136 AICc=-118
resl1313 <- residuals(TFRml1313)
acf(resl1313, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,3) for TFR with log transformation")
pacf(resl1313, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,3) for TFR with log transformation")
arimal1313 <- forecast(TFRml1313, h = 12)
predl1313 <- arimal1313$mean
msel1313 <- mean((actualTFR - exp(predl1313))^2) # 0.02602064
absl1313 <- mean(abs(actualTFR - exp(predl1313))) # 0.121592
```

\newpage
```{r page-1l TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1214 <- Arima(log(TFR_ts), order=c(12,1,4))
TFRml1214 # AIC=-131.83 AICc=-113.83
resl1214 <- residuals(TFRml1214)
acf(resl1214, lag.max = 50, main = "ACF of Residuals for ARIMA(12,1,4) for TFR with log transformation")
pacf(resl1214, lag.max = 50, main = "PACF of Residuals for ARIMA(12,1,4) for TFR with log transformation")
arimal1214 <- forecast(TFRml1214, h = 12)
predl1214 <- arimal1214$mean
msel1214 <- mean((actualTFR - exp(predl1214))^2) # 0.02101819
absl1214 <- mean(abs(actualTFR - exp(predl1214))) # 0.1132223
```

\newpage
```{r page-1m TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1315 <- Arima(log(TFR_ts), order=c(13,1,5))
TFRml1315 # AIC=-133.96 AICc=-110.21
resl1315 <- residuals(TFRml1315)
acf(resl1315, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,5) for TFR with log transformation")
pacf(resl1315, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,5) for TFR with log transformation")
arimal1315 <- forecast(TFRml1315, h = 12)
predl1315 <- arimal1315$mean
msel1315 <- mean((actualTFR - exp(predl1315))^2) # 0.05826379
absl1315 <- mean(abs(actualTFR - exp(predl1315))) # 0.1840232
```

\newpage
```{r page-1n TFR log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1215 <- Arima(log(TFR_ts), order=c(12,1,5))
TFRml1215 # AIC=-130.02 AICc=-109.3
resl1215 <- residuals(TFRml1215)
acf(resl1215, lag.max = 50, main = "ACF of Residuals for ARIMA(12,1,5) for TFR with log transformation")
pacf(resl1215, lag.max = 50, main = "PACF of Residuals for ARIMA(12,1,5) for TFR with log transformation")
arimal1215 <- forecast(TFRml1215, h = 12)
predl1215 <- arimal1215$mean
msel1215 <- mean((actualTFR - exp(predl1215))^2) # 0.02172126
absl1215 <- mean(abs(actualTFR - exp(predl1215))) # 0.1147922
```

\newpage
```{r page-1o TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))),
     type = 'l',
     main = "Plot of Differenced Twice Log Transformation for Total Fertility Rate (TFR)")
unitroot_kpss(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))))
BirthAndFertility |>
  ACF(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Twice Differenced Log Transformation of Total Fertility Rate (TFR)")
BirthAndFertility |>
  PACF(diff(diff(log(TFR$`Total Fertility Rate (TFR)`))), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Twice Differenced Log Transformation of Total Fertility Rate (TFR)")
TFRml1520 <- Arima(log(TFR_ts), order=c(15,2,0))
TFRml1520 # AIC=-133.67 AICc=-117.67
resl1520 <- residuals(TFRml1520)
acf(resl1520, lag.max = 50, main = "ACF of Residuals for ARIMA(15,2,0) for TFR with log transformation")
pacf(resl1520, lag.max = 50, main = "PACF of Residuals for ARIMA(15,2,0) for TFR with log transformation")
arimal1520 <- forecast(TFRml1520, h = 12)
predl1520 <- arimal1520$mean
msel1520 <- mean((actualTFR - exp(predl1520))^2) # 0.01869052
absl1520 <- mean(abs(actualTFR - exp(predl1520))) # 0.1056717
```
\newpage
```{r page-1p TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1522 <- Arima(log(TFR_ts), order=c(15,2,2))
TFRml1522 # AIC=-135.54 AICc=-114.17
resl1522 <- residuals(TFRml1522)
acf(resl1522, lag.max = 50, main = "ACF of Residuals for ARIMA(15,2,2) for TFR with log transformation")
pacf(resl1522, lag.max = 50, main = "PACF of Residuals for ARIMA(15,2,2) for TFR with log transformation")
arimal1522 <- forecast(TFRml1522, h = 12)
predl1522 <- arimal1522$mean
msel1522 <- mean((actualTFR - exp(predl1522))^2) # 0.02235162
absl1522 <- mean(abs(actualTFR - exp(predl1522))) # 0.1187815
```

\newpage
```{r page-1q TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1422 <- Arima(log(TFR_ts), order=c(14,2,2))
TFRml1422 # AIC=-130.49 AICc=-111.94
resl1422 <- residuals(TFRml1422)
acf(resl1422, lag.max = 50, main = "ACF of Residuals for ARIMA(14,2,2) for TFR with log transformation")
pacf(resl1422, lag.max = 50, main = "PACF of Residuals for ARIMA(14,2,2) for TFR with log transformation")
arimal1422 <- forecast(TFRml1422, h = 12)
predl1422 <- arimal1422$mean
msel1422 <- mean((actualTFR - exp(predl1422))^2) # 0.01087075
absl1422 <- mean(abs(actualTFR - exp(predl1422))) # 0.08352761
```

\newpage
```{r page-1r TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1322 <- Arima(log(TFR_ts), order=c(13,2,2))
TFRml1322 # AIC=-132.39 AICc=-116.39
resl1322 <- residuals(TFRml1322)
acf(resl1322, lag.max = 50, main = "ACF of Residuals for ARIMA(13,2,2) for TFR with log transformation")
pacf(resl1322, lag.max = 50, main = "PACF of Residuals for ARIMA(13,2,2) for TFR with log transformation")
arimal1322 <- forecast(TFRml1322, h = 12)
predl1322 <- arimal1322$mean
msel1322 <- mean((actualTFR - exp(predl1322))^2) # 0.007426324
absl1322 <- mean(abs(actualTFR - exp(predl1322))) # 0.06826933
```

\newpage
```{r page-1s TFR log 2nd diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TFRml1222 <- Arima(log(TFR_ts), order=c(12,2,2))
TFRml1222 # AIC=-127.34 AICc=-113.62
resl1222 <- residuals(TFRml1222)
acf(resl1222, lag.max = 50, main = "ACF of Residuals for ARIMA(12,2,2) for TFR with log transformation")
pacf(resl1222, lag.max = 50, main = "PACF of Residuals for ARIMA(12,2,2) for TFR with log transformation")
arimal1222 <- forecast(TFRml1222, h = 12)
predl1222 <- arimal1222$mean
msel1222 <- mean((actualTFR - exp(predl1222))^2) # 0.0498791
absl1222 <- mean(abs(actualTFR - exp(predl1222))) # 0.1720094
```
The ACF for TFR shows significant positive autocorrelations from lags 1 to 9, while TLB shows significant positive autocorrelations from lags 1 to 6, with a slow decrease in the ACF as the lag increases. This suggests that neither series is white noise, as there are multiple spikes outside the significance bounds.

The presence of a strong trend and slow decrease confirms that both TFR and TLB are non-stationary series. The absence of a sharp cut-off suggests that differencing or an autoregressive (AR) model may be suitable for this series.
An extended ACF plot for TFR confirms that significant positive autocorrelations persist up to lag 10.

\newpage
```{r page-2 TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
autoplot(BirthAndFertility,`Total Live-Births`)+
  labs(title = "Initial Plot of Total Live-Births")
autoplot(TLB) +
  labs(title = "Initial Plot of Total Live-Births (TLB) until 2012")
```

From the initial plots of TFR and TLB, both series show a long-term decreasing trend over time, with TFR exhibiting a more pronounced decline. A notable change is observed around 1980 to 1990, where both series display a temporary broad peak before continuing their downward trend. As the data is yearly, seasonality cannot be determined. While both series may suggest cyclical behaviour, the trend would need to be removed to confirm this.

```{r page-2b TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(diff(TLB$`Total Live-Births`),
     type = 'l',
     main = "Plot of Differenced Total Live Births (TLB) until 2012")
unitroot_kpss(diff(TLB$`Total Live-Births`)) # stationary
BirthAndFertility |>
  ACF(diff(TLB$`Total Live-Births`), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Total Live Births (TLB)")
BirthAndFertility |>
  PACF(diff(TLB$`Total Live-Births`), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Total Live Births (TLB)")
TLBm1410 <- Arima(TLB_ts, order=c(14,1,0))
TLBm1410 # AIC=975.33 AICc=988.66
resb1410 <- residuals(TLBm1410)
acf(resb1410, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,0) for TLB")
pacf(resb1410, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,0) for TLB")
arima1410 <- forecast(TLBm1410, h = 12)
pred1410 <- arima1410$mean
mse1410 <- mean((actualTLB - pred1410)^2) # 67367429
abs1410 <- mean(abs(actualTLB - pred1410)) # 6853.008
```

\newpage
```{r page-2c TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBm1411 <- Arima(TLB_ts, order=c(14,1,1))
TLBm1411 # AIC=975.62 AICc=991.17
resb1411 <- residuals(TLBm1411)
acf(resb1411, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,1) for TLB")
pacf(resb1411, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,1) for TLB")
arima1411 <- forecast(TLBm1411, h = 12)
pred1411 <- arima1411$mean
mse1411 <- mean((actualTLB - pred1411)^2) # 84179161
abs1411 <- mean(abs(actualTLB - pred1411)) # 7769.997
```

\newpage
```{r page-2d TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBm1311 <- Arima(TLB_ts, order=c(13,1,1))
TLBm1311 # AIC=974.57 AICc=987.9
resb1311 <- residuals(TLBm1311)
acf(resb1311, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,1) for TLB")
pacf(resb1311, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,1) for TLB")
arima1311 <- forecast(TLBm1311, h = 12)
pred1311 <- arima1311$mean
mse1311 <- mean((actualTLB - pred1311)^2) # 82856991
abs1311 <- mean(abs(actualTLB - pred1311)) # 7698.072
```

\newpage
```{r page-2e TLB, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBm1312 <- Arima(TLB_ts, order=c(13,1,2))
TLBm1312 # AIC=973.01 AICc=988.56
resb1312 <- residuals(TLBm1312)
acf(resb1312, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,2) for TLB")
pacf(resb1312, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,2) for TLB")
arima1312 <- forecast(TLBm1312, h = 12)
pred1312 <- arima1312$mean
mse1312 <- mean((actualTLB - pred1312)^2) # 54382525
abs1312 <- mean(abs(actualTLB - pred1312)) # 6014.802
```

\newpage
```{r page-2f TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
plot(log(TLB$`Total Live-Births`),
     type = 'l',
     main = "Plot of Log Transformation for Total Live-Births (TLB) until 2012")
plot(diff(log(TLB$`Total Live-Births`)),
     type = 'l',
     main = "Plot of Differenced Log Transformation for Total Live-Births (TLB) until 2012")
unitroot_kpss(diff(log(TLB$`Total Live-Births`))) # stationary
BirthAndFertility |>
  ACF(diff(log(TLB$`Total Live-Births`)), lag_max = 50) |>
  autoplot() +
  labs(title = "ACF Plot for Differenced Log Transformation of Total Live-Births (TLB)")
BirthAndFertility |>
  PACF(diff(log(TLB$`Total Live-Births`)), lag_max = 50) |>
  autoplot() +
  labs(title = "PACF Plot for Differenced Log Transformation of Total Live-Births (TLB)")
TLBml1510 <- Arima(log(TLB_ts), order=c(15,1,0))
TLBml1510 # AIC=-145.56 AICc=-130.02
resl1510 <- residuals(TLBml1510)
acf(resl1510, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,0) for TLB with log transformation")
pacf(resl1510, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,0) for TLB with log transformation")
arimal1510 <- forecast(TLBml1510, h = 12)
predl1510 <- arimal1510$mean
msel1510 <- mean((actualTFR - exp(predl1510))^2) # 2239721240
absl1510 <- mean(abs(actualTFR - exp(predl1510))) # 47245.34
```
\newpage
```{r page-2g TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBml1511 <- Arima(log(TLB_ts), order=c(15,1,1))
TLBml1511 # AIC=-145.43 AICc=-127.43
resl1511 <- residuals(TLBml1511)
acf(resl1511, lag.max = 50, main = "ACF of Residuals for ARIMA(15,1,1) for TLB with log transformation")
pacf(resl1511, lag.max = 50, main = "PACF of Residuals for ARIMA(15,1,1) for TLB with log transformation")
arimal1511 <- forecast(TLBml1511, h = 12)
predl1511 <- arimal1511$mean
msel1511 <- mean((actualTFR - exp(predl1511))^2) # 2173041601
absl1511 <- mean(abs(actualTFR - exp(predl1511))) # 46542.86
```

\newpage
```{r page-2h TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBml1411 <- Arima(log(TLB_ts), order=c(14,1,1))
TLBml1411 # AIC=-140.56 AICc=-125.02
resl1411 <- residuals(TLBml1411)
acf(resl1411, lag.max = 50, main = "ACF of Residuals for ARIMA(14,1,1) for TLB with log transformation")
pacf(resl1411, lag.max = 50, main = "PACF of Residuals for ARIMA(14,1,1) for TLB with log transformation")
arimal1411 <- forecast(TLBml1411, h = 12)
predl1411 <- arimal1411$mean
msel1411 <- mean((actualTFR - exp(predl1411))^2) # 2173041601
absl1411 <- mean(abs(actualTFR - exp(predl1411))) # 46542.86
```

\newpage
```{r page-2i TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBml1311 <- Arima(log(TLB_ts), order=c(13,1,1))
TLBml1311 # AIC=-141.45 AICc=-128.11
resl1311 <- residuals(TLBml1311)
acf(resl1311, lag.max = 50, main = "ACF of Residuals for ARIMA(13,1,1) for TLB with log transformation")
pacf(resl1311, lag.max = 50, main = "PACF of Residuals for ARIMA(13,1,1) for TLB with log transformation")
arimal1311 <- forecast(TLBml1311, h = 12)
predl1311 <- arimal1311$mean
msel1311 <- mean((actualTFR - exp(predl1311))^2) # 2200303964
absl1311 <- mean(abs(actualTFR - exp(predl1311))) # 46836.59
```

\newpage
```{r page-2j TLB log diff, echo = FALSE, warning = FALSE, message = FALSE, out.width = "70%", fig.align = 'center'}
TLBml1213 <- Arima(log(TLB_ts), order=c(12,1,3))
TLBml1213 # AIC=-139.69 AICc=-124.15
resl1213 <- residuals(TLBml1213)
acf(resl1213, lag.max = 50, main = "ACF of Residuals for ARIMA(12,1,3) for TLB with log transformation")
pacf(resl1213, lag.max = 50, main = "PACF of Residuals for ARIMA(12,1,3) for TLB with log transformation")
arimal1213 <- forecast(TLBml1213, h = 12)
predl1213 <- arimal1213$mean
msel1213 <- mean((actualTFR - exp(predl1213))^2) # 1752591080
absl1213 <- mean(abs(actualTFR - exp(predl1213))) # 41851.92
```

\newpage
# Results
```{r echo = FALSE}
# 5/30 points
# flawlessly estimated relevant time series models(s) and/or forecasts in R

# effectively and appropriately used composition of time series methods and advanced modelling for data

# critically interpreted and evaluated model fits with appropriate comprehensive use of diagnostics
```

| TFR Model | AIC  | AICc | MSE | MAE |
|--------|------|------|------|------| 
| ARIMA(16,1,0) | -63.99 | -45.99 | 0.04753075 | 0.1690442 |**
| ARIMA(16,1,1) | -62.39 | -42.66 | 0.04783471 | 0.1700046 |
| ARIMA(15,1,1) | -60.26 | -42.26 | 0.08431368 | 0.223524 |
| ARIMA(14,1,3) | -60.99 | -40.26 | 0.01899351 | 0.09964934 |
| ARIMA(12,2,3) | -62.57 | -46.57 | 0.01218546 | 0.08274547 |**
| ARIMA(11,2,3) | -52.91 | -39.19 | 0.05260825 | 0.1823268 |
| ARIMA(10,2,3) | -54.69 | -43.02 | 0.05680021 | 0.1898403 |



| Log Transformed TFR Model | AIC  | AICc | MSE | MAE |
|--------|------|------|------|------| 
| ARIMA(13,1,3) | -136 | -118 | 0.02602064 | 0.121592 | **
| ARIMA(12,1,4) | -131.83 | -113.83 | 0.02101819 | 0.1132223 |
| ARIMA(13,1,5) | -133.96 | -110.21 | 0.05826379 | 0.1840232 |
| ARIMA(12,1,5) | -130.02 | -109.3 | 0.02172126 | 0.1147922 |
| ARIMA(15,2,0) | -133.67 | -117.67 | 0.01869052 | 0.1056717 |
| ARIMA(15,2,2) | -135.54 | -114.17 | 0.02235162 | 0.1187815 |
| ARIMA(14,2,2) | -130.49 | -111.94 | 0.01087075 | 0.08352761 |
| ARIMA(13,2,2) | -132.39 | -116.39 | 0.007426324 | 0.06826933 |
| ARIMA(12,2,2) | -127.34 | -113.62 | 0.0498791 | 0.1720094 |

| TLB Model | AIC  | AICc | MSE | MAE |
|--------|------|------|------|------|
| ARIMA(14,1,0) | 975.33 | 988.66 | 67367429 | 6853.008 |
| ARIMA(14,1,1) | 975.62 | 991.17 | 84179161 | 7769.997 |
| ARIMA(13,1,1) | 974.57 | 987.9 | 82856991 | 7698.072 | **
| ARIMA(13,1,2) | 973.01 | 988.56 | 54382525 | 6014.802 | **

| Log Transformed TLB Model | AIC  | AICc | MSE | MAE |
|--------|------|------|------|------| 
| ARIMA(15,1,0) | -145.56 | -130.02 | 2239721240 | 47245.34 | **
| ARIMA(15,1,1) | -145.43 | -127.43 | 2173041601 | 46542.86 |
| ARIMA(14,1,1) | -140.56 | -125.02 | 2211919788 | 46958.53 |
| ARIMA(13,1,1) | -141.45 | -128.11 | 2200303964 | 46836.59 |
| ARIMA(12,1,3) | -139.69 | -124.15 | 1752591080 | 41851.92 |

# Discussion
```{r echo = FALSE}
# 5/30 points
# thoroughly integrated limitations of models and processes into statistical conclusions

# comprehensively evaluated impact of model limitations on answering research question
```

# Conclusion
```{r echo = FALSE}
# 
```

```{r echo = FALSE}
# report presentation 2/30 points
# report structured with appropriate length, use of sections and highly professional style

# written with precise, concise, and correct scientific language with flawless grammar and professional tone

# visualisation presented with professional quality
```

# References
```{r echo = FALSE}
# 5/30 points
# extension relevant literature reference, all correctly cited and referenced to in the main report

# appendix well presented as a coherent follow up to the main report
```

# Appendices / Statisical Appendix
```{r echo = FALSE}
# 
```

\newpage
# Research Questions
Can the ARIMA model accurately forecast the Singapore's Total Live Births (TLB) and Total Fertility Rate (TFR) from 2013 to 2024 given the data from 1960 to 2012, capturing the short-term fluctuations? 

# Statistical Appendix
The following statistical methods were applied throughout the analysis in the order presented below.

## Autoregressive Model (AR)
$$
y_t = c + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \cdots + \phi_p y_{t-p} + \varepsilon_t
$$
where:

\( y_t \) is the value at time t

\( \phi_i \) are model parameters

\( \varepsilon_t \) is white noise

Models the current values using its previous values (lags)

## Residuals
$$
e_t = y_t - \hat{y}_t
$$
\( y_t \) is the actual observed value at time t

\( \hat{y}_t \) is the predicted value at time t

The error which is the difference between the actual observed value and predicted value at time t, used to check the chosen model accounts for all the patterns

## Autocorrelation Function (ACF)
$$
\rho_k = \frac{\text{Cov}(y_t, y_{t-k})}{\text{Var}(y_t)}
$$
where:

\( \rho_k \) is the autocorrelation at lag \( k \)

Measures the correlation between a series \( y_t \) and lagged values \( y_{t-k} \), which checks for trend, seasonality and whether residuals are white noise

## Box-Cox Transformation
$$
y^{(\lambda)} =
\begin{cases}
\frac{y^\lambda - 1}{\lambda}, & \lambda \ne 0 \\
\log(y), & \lambda = 0
\end{cases}
$$

\(\lambda\) was estimated using the Guerrero method

## STL Decomposition
$$
y_t = T_t + R_t
$$
where:

\( T_t \) is the trend component,

\( R_t \) is the remainder

Decomposes time series into trend and remainder components, usually with the seasonality component as well but the data in this analysis is yearly

## Moving Average (MA)
$$
MA_k = \frac{1}{k} \sum_{i=-(k-1)/2}^{(k-1)/2} y_{t+i}
$$
k is the window size

In this analysis when k = 5, the moving average at time t is calculated using the 2 observations before and 2 observations after t. This process smooths out short-term fluctuations in the series

<https://github.com/twalepear/TSA.git>
